Learning Auditory Similarities Between Instrumental Playing Techniques

### 2020-05-16 paper/paper.tex
cite schorkhuber and klapuri, write more on scattering

\subsection*{Outline}
The \lnameref{sec:soa} section reviews recent research in the computational modeling of musical timbre.
The \lnameref{sec:survey} section describes our computer--human interface for annotating auditory similarities across playing techniques.
The \lnameref{sec:methods} section presents the technical components of our machine listening system, namely joint time--frequency scattering and large-margin nearest neighbors.
The \lnameref{sec:results} section describes our experimental benchmark.
The \lnameref{sec:discussion} section conducts an ablation study of our best performing system.
The \lnameref{sec:conclusion} summarizes the implications of our findings for future research in music information retrieval.\footnote{For the sake of research reproducibility, the source code for the experimental protocol of this paper is available online, alongside anonymized data from human subjects:  \url{https://github.com/mathieulagrange/paperPlayingTechniquesSimilarityLearning}} %TODO: simplify this URL to lostanlen2020jasmp

\section*{\nmu Related work}
\label{sec:soa}

@@

Convolving the wavelets in this filterbank with an input waveform $\boldsymbol{x}\in\mathbf{L}^2(\mathbb{R}$, followed by an application of the pointwise complex modulus, yields the wavelet scalogram
\begin{equation}
\mathbf{U_1}\boldsymbol{x}(t,\log_2\lambda) =
\big\vert
\boldsymbol{x}
\ast
\boldsymbol{\psi}_{\lambda}
\big\vert(t)
=
\left\vert
\int_{\mathbb{R}}
\boldsymbol{x}(t^{\prime})
\,
\boldsymbol{\psi}_{\lambda} (t - t^{\prime})
\;
\mathrm{d}{t^\prime}
\right\vert,
\end{equation}
indexed by time $t$ and log-frequency $\log_2\lambda$.
Then, we define a two-dimensional Morlet wavelet $\boldsymbol{\Psi}$, taking a spectrotemporal pair $(t, \log_2 \lambda)$ as input:
\begin{equation}
\Psi : (t,\log_2 \lambda) \longmapsto 
\exp\left(-\dfrac{t^2+(\log_2 \lambda)^2}{2\sigma_{\Psi}^2}\right) 
\Big[ \exp\big(2\pi \mathrm{i} (t + \log_2 \lambda)\big) - \kappa_{\Psi} \Big].
\end{equation}
In the equation above, the Gaussian width $\sigma_{\Psi}$ is chosen so that the quality factor of the Morlet wavelet $\Psi$ is equal to $1$, both over the time dimension and over the log-frequency dimension.
Furthermore, the corrective term $\kappa_{\Psi}$ ensures that $\boldsymbol{\Psi}$ has a null average over $\mathbb{R}^2$, similarly to Equation \ref{eq:psi}.
Following the terminology of spectrotemporal receptive fields (STRF, see \lnameref{sec:soa} section), we define a two-dimensional wavelet filterbank of the form:
\begin{equation}
\mathbf{\Psi}_{\alpha,\beta} : (t, \log_2 \lambda) \longmapsto
\mathbf{\Psi}\big(\alpha t, \beta \log_2 \lambda \big),
\end{equation}
where $\alpha$ is a temporal modulation rate and $\beta$ is a temporal modulation scale.

While $\alpha$ is measured in Hertz and is strictly positive, $\beta$ is measured in cycles per octaves and may take positive as well as negative values.
Both $\alpha$ and $\beta$ are discretized by geometric progressions of common ratio equal to $2$.
Furthermore, the edge case $\beta=0$ corresponds to $\mathbf{\Psi}_{\alpha,\beta}$ being a Gaussian low-pass filter over the log-frequency dimension, while remaining a Morlet band-pass filter of center frequency $\alpha$ over the time dimension.

% TODO define U2.
% Briefly define S2.

@@

Over our cohort of $K=31$ subjects, separable time--frequency scattering achieves a precision at rank five (P@5) of $\textbf{91.9\%} \pm 4$. This figure is noticeably worse than joint time--frequency scattering ($99.0\% \pm 1$), all other things being equal.
This gap in performance, together with the theory of spectrotemporal receptive fields (STRF) in auditory neuroscience (see \lnameref{sec:soa} section), demonstrates the importance of joint spectrotemporal modulations in the modeling of timbre similarity across instrumental playing techniques.

>>>>

\subsection*{Outline}
The \lnameref{sec:related-work} section reviews recent research in the computational modeling of musical timbre.
The \lnameref{sec:survey} section describes our computer--human interface for annotating auditory similarities across playing techniques.
The \lnameref{sec:methods} section presents the technical components of our machine listening system, namely joint time--frequency scattering and large-margin nearest neighbors.
The \lnameref{sec:results} section describes our experimental benchmark.
The \lnameref{sec:discussion} section conducts an ablation study of our best performing system.
The \lnameref{sec:conclusion} summarizes the implications of our findings for future research in music information retrieval.\footnote{For the sake of research reproducibility, the source code for the experimental protocol of this paper is available online, alongside anonymized data from human subjects:  \url{https://github.com/mathieulagrange/paperPlayingTechniquesSimilarityLearning}} %TODO: simplify this URL to lostanlen2020jasmp

\section*{\nmu Related work}
\label{sec:related-work}

@@

Convolving the wavelets in this filterbank with an input waveform $\boldsymbol{x}\in\mathbf{L}^2(\mathbb{R})$, followed by an application of the pointwise complex modulus, yields the wavelet scalogram
\begin{equation}
\mathbf{U_1}\boldsymbol{x}(t,\lambda) =
\big\vert
\boldsymbol{x}
\ast
\boldsymbol{\psi}_{\lambda}
\big\vert(t)
=
\left\vert
\int_{\mathbb{R}}
\boldsymbol{x}(t^{\prime})
\,
\boldsymbol{\psi}_{\lambda} (t - t^{\prime})
\;
\mathrm{d}{t^\prime}
\right\vert,
\end{equation}
which is discretized according to a dyadic multirate scheme, similarly to the constant-$Q$ transform of \cite{schorkhuber2010smc}.
Then, we define a two-dimensional Morlet wavelet $\boldsymbol{\Psi}$ of the form
\begin{equation}
\Psi : (t, u) \longmapsto 
\exp\left(-\dfrac{t^2+u^2}{2\sigma_{\Psi}^2}\right) 
\Big[ \exp\big(2\pi \mathrm{i} (t + u)\big) - \kappa_{\Psi} \Big],
\end{equation}
taking two real variables $t$ and $u$ as input.
The former is the time variable while the latter is the base-two logarithm of frequency: $u = \log_2 \lambda$.
Note that $u$ roughly corresponds to the human perception of relative pitch \cite{lostanlen2020icassp}.

In the equation above, the Gaussian width $\sigma_{\Psi}$ is chosen so that the quality factor of the Morlet wavelet $\Psi$ is equal to $1$, both over the time dimension and over the log-frequency dimension.
Furthermore, the corrective term $\kappa_{\Psi}$ ensures that $\boldsymbol{\Psi}$ has a null average over $\mathbb{R}^2$, similarly to Equation \ref{eq:psi}.
From $\mathbf{\Psi}$, we define a two-dimensional wavelet filterbank of the form:
\begin{equation}
\mathbf{\Psi}_{\alpha,\beta} : (t, \log_2 \lambda) \longmapsto
\alpha \, \beta \, \mathbf{\Psi}(\alpha t, \beta \log_2 \lambda).
\end{equation}
In the equation above, $\alpha$ is a temporal modulation rate and $\beta$ is a temporal modulation scale, following the terminology of spectrotemporal receptive fields (STRF, see \lnameref{sec:related-work} section).
While $\alpha$ is measured in Hertz and is strictly positive, $\beta$ is measured in cycles per octaves and may take positive as well as negative values.
Both $\alpha$ and $\beta$ are discretized by geometric progressions of common ratio equal to $2$.
Furthermore, the edge case $\beta=0$ corresponds to $\mathbf{\Psi}_{\alpha,\beta}$ being a Gaussian low-pass filter over the log-frequency dimension, while remaining a Morlet band-pass filter of center frequency $\alpha$ over the time dimension.

\begin{equation}
\mathbf{U_2}\boldsymbol{x} (t, \lambda, \alpha, \beta) =
\left\vert
\iint_{\mathbb{R}^2}
\mathrm{d}t^\prime \, \mathrm{d}u^\prime
\right\vert
\end{equation}

@@

Over our cohort of $K=31$ subjects, separable time--frequency scattering achieves a precision at rank five (P@5) of $\textbf{91.9\%} \pm 4$. This figure is noticeably worse than joint time--frequency scattering ($99.0\% \pm 1$), all other things being equal.
This gap in performance, together with the theory of spectrotemporal receptive fields (STRF) in auditory neuroscience (see \lnameref{sec:related-work} section), demonstrates the importance of joint spectrotemporal modulations in the modeling of timbre similarity across instrumental playing techniques.
